# Exploratory Data Analysis with R
## By Jason Zmudka

========================================================


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.


library(ggplot2)
library(knitr)
library(RColorBrewer)
library(gridExtra)
library(dplyr)
```
  
 This report analyzes a data set of 4898 white wines. Each wine's quality was 
 rated by a taster on a scale of 1 to 10. There are also 11 different
 chemical properties that were measured.


```{r,  echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
# Load the Data
wines <- read.csv('wineQualityWhites.csv')
str(wines)
```
```{r,  echo=FALSE,message=FALSE, warning=FALSE, summary}
summary(wines)
```

In the data, there are 4898 separate observations and 12 Variables. 

# Univariate Plots Section

```{r echo=FALSE, Univariate_Plots}
ggplot(aes(x = quality),data = wines)+
  geom_histogram(binwidth = 1, color = 'black', fill = 'blue')
```
```{r  echo=FALSE,message=FALSE, warning=FALSE }
summary(wines$quality)
```
The quality factor gives us this normal looking histogram. The majority of wines
have a quality rating of 5 or 6.



```{r echo=FALSE, message = FALSE, alcohol}
ggplot(aes(x = alcohol),data = wines)+
  geom_histogram(binwidth = 0.5, color = 'black', fill = 'orange')
```
```{r echo=FALSE,message=FALSE, warning=FALSE }
summary(wines$alcohol)

```
The mean alcohol level on the white wines is about 10.5 but some have much
higher levels. The histogram is positively skewed.
What other properties do the wines with high alcohol content have?

```{r  echo=FALSE,message=FALSE, warning=FALSE}
# top wines by alcohol content
head(wines[order(-wines$alcohol),])
```

It looks like they are mostly above average quality

```{r echo=FALSE, message=FALSE, residual.sugar}
ggplot(aes(x = residual.sugar),data = wines)+
  geom_histogram(color = 'black', fill = 'green')

```

```{r  echo=FALSE,message=FALSE, warning=FALSE}
summary(wines$residual.sugar)
```
When looking at the residual sugar data, it was skewed badly to the left,
long tailed on the right and there were some outliers.

```{r echo=FALSE, message=FALSE, warning=FALSE, residual.sugar.log10}
ggplot(aes(x = residual.sugar),data = subset(wines, residual.sugar<=30))+
  geom_histogram(color = 'black', fill = 'green')+
  scale_y_log10()
```


Here I discarded the measurements above 30 and adjusted the Y scale with log10
to get a more normal visualization. Apparently most white wines are low on sugar
but some are really high.

```{r echo=FALSE, message=FALSE, volatile.acidity}
p1 <- ggplot(aes(x = volatile.acidity),data = wines)+
  geom_histogram(binwidth = 0.01, color = 'black', fill = 'yellow')
p2 <- ggplot(aes(x = fixed.acidity),data = wines)+
  geom_histogram(binwidth = 0.1, color = 'black', fill = 'blue')
grid.arrange(p1,p2,ncol=1)

```


Both the volatile acidity and the fixed acidity give us similar distributions.
I wonder if they have a relationship. I'll have to study this more later.

```{r echo=FALSE, message=FALSE, density}
ggplot(aes(x = density),data = wines)+
  geom_histogram(color = 'black', fill = 'brown')
```
```{r  echo=FALSE,message=FALSE, warning=FALSE}
summary(wines$density)
```

Density gives us a normal looking distribution with a couple outliers.
I wanted to see what the properties the outliers have.


```{r echo=FALSE,message=FALSE, warning=FALSE }
# top wines by density
head(wines[order(-wines$density),])


```

They all have high levels of residual sugar. Looks like there may be a
relationship.


```{r echo=FALSE, message=FALSE, sulphates}
ggplot(aes(x = sulphates),data = wines)+
  geom_histogram(binwidth = 0.05, color = 'black', fill = 'orange')
```

Sulphates give us a fairly normal distribution with a long tail.

```{r echo=FALSE, message=FALSE, pH}
ggplot(aes(x = pH),data = wines)+
  geom_histogram(binwidth = 0.05, color = 'black', fill = 'blue')
```

When I looked at the pH data, I got this very normal looking histogram.


# Univariate Analysis


### What is the structure of your dataset?

The data consists of 4898 rows and 12 features:

1. fixed acidity (tartaric acid - g / dm^3) <br />
2. volatile acidity (acetic acid - g / dm^3) <br />
3. citric acid (g / dm^3) <br />
4. residual sugar (g / dm^3) <br />
5. chlorides (sodium chloride - g / dm^3 <br />
6. free sulfur dioxide (mg / dm^3) <br />
7. total sulfur dioxide (mg / dm^3) <br />
8. density (g / cm^3) <br />
9. pH <br />
10. sulphates (potassium sulphate - g / dm3) <br />
11. alcohol (% by volume) <br />
12. quality (score between 0 and 10) <br />


### What is/are the main feature(s) of interest in your dataset?

Quality is the main feature of the data. It has  simple 1-10 rating system.
I plan on analyzing the other 11 columns to see how they affect quality level.

<h3> What other features in the dataset do you think will help support your
investigation into your feature(s) of interest. </h3>

I think residual sugar levels and alcohol levels will play a big role. I also 
can see where having high levels of some characteristics like sulfates and
acidity might affect the quality level negatively.

### Did you create any new variables from existing variables in the dataset?

No, the measurements given were very conventional.

<h3>  Of the features you investigated, were there any unusual distributions?
Did you perform any operations on the data to tidy, adjust, or change the
form of the data? If so, why did you do this? </h3>

The distribution of residual.sugar was highly skewed to the left and had a very
long tail. The vast majority of the wines had low sugar but there were some
that had higher levels of sugar. I scaled the y variable with log10,
and the histogram showed me that there was consistent amount of wines with
higher residual sugar.


# Bivariate Plots Section


```{r, echo=FALSE,correlation.matrix}
cor.matrix <- cor(wines)
round(cor.matrix,2)
```

Looking at the correlation table, there are a few things that stand out.
Density strongly correlates with residual sugar and alcohol. Free sulfur dioxide
strongly correlates with total sulfur dioxide. Quality, our most important
feature, most strongly correlates with alcohol content.


```{r echo = FALSE, message = FALSE, warning = FALSE}

ggplot(aes(x = quality, y = alcohol), data = wines)+
  geom_point(alpha = 0.1)+
  geom_smooth(color = 'red', method = "lm", se = FALSE)

```


Here we can see the relationship between quality and alcohol content. There are
few white wines with a high rating that are low alcohol. The red line is the 
line of regression.



```{r echo = FALSE, message = FALSE, warning = FALSE}
ggplot(aes(x = density, y = residual.sugar), data = wines)+
  geom_point(size = 2)+
  ylim(0,25)+
  xlim(0.985, 1.005)


```

Looking at this scatter plot, it appears that having a high sugar content makes 
for a more dense wine.

```{r echo=FALSE, message=FALSE, warning = FALSE}

ggplot(aes(x = density, y = alcohol), data = wines)+
  geom_point(size = 2)+
  xlim(0.985, 1.005)

```

In this scatter plot, we see that the higher the alcohol content, the lower
the density.



```{r echo=FALSE, message=FALSE, warning = FALSE}
ggplot(aes(x = alcohol, y = residual.sugar), data = wines)+
  geom_point(size = 2)+
  ylim(0,25)+
  geom_smooth(color = 'red', method = "lm", se = FALSE)
```

As I suspected guessed, alcohol and residual sugar are inversely correlated.
The red line is the line of regression.

```{r echo = FALSE, message = FALSE, warning = FALSE}
ggplot(aes(x = quality, y = residual.sugar), data = wines)+
  geom_jitter(alpha = 0.1, size=2)+
  geom_smooth(color = 'red', method = "lm", se = FALSE)+
  ylim(0,30)

```

I used jitter and alpha to show that as expected, there are fewer high quality 
white wines with high residual sugar. However, there are some white wines with 
high residual sugar. The red line is the line of regression.

```{r echo=FALSE, message=FALSE, warning = FALSE}
ggplot(aes(x = fixed.acidity, y = volatile.acidity), data = wines)+
  geom_point(size = 2)+
  xlim(4,12)+
  ylim(0,0.6)

```

This one surprised me.
Although they have similar histogram plots, there is no obvious relationship
between fixed acidity and volatile acidity.


```{r echo=FALSE, message=FALSE, warning = FALSE}

# sorted total sulfur dioxide into buckets
wines$total.sulfur.dioxide.bucket <- cut(wines$total.sulfur.dioxide,
                                         c(0,50,100,150,200,250,300))

ggplot(aes(x=round(total.sulfur.dioxide/5) * 5, y=quality), data=subset(wines,                                                           !is.na(total.sulfur.dioxide.bucket)))+
  geom_line(stat = 'summary', fun = 'mean', linetype = 1)+
  geom_smooth()+
  xlim(0,300)
  

```

I wondered if some factors had an ideal range of values where the quality of
the wine would be the highest. Here I sorted total sulfur dioxide values into
buckets and plotted the conditional means of the quality factor. As you can see,
the sweet spot for total sulfur dioxide is just around 100.


```{r echo=FALSE, message=FALSE, warning = FALSE}
ggplot(aes(x = factor(quality), y = volatile.acidity), data = wines)+
  geom_boxplot()
```

In the description of the data, it says that high levels of volatile acidity
would lead to a more vinegar like taste. This box plot shows that high levels
of it are definitely less desirable in terms of quality.

# Bivariate Analysis


<h3> Talk about some of the relationships you observed in this part of the 
investigation. How did the feature(s) of interest vary with other features in 
the dataset? </h3>

Quality has a strongly correlated relationship to alcohol. Many of other
factors were inversely correlated to quality indicating that you wouldn't
want their levels to be too high. I found that total.sulfur.dioxide had
a range where it was best.

<h3> Did you observe any interesting relationships between the other features
(not the main feature(s) of interest)? </h3>

The relationships between residual sugar, alcohol, and density were interesting.
Alcohol is less dense than water while adding sugar to water adds density.
During the fermentation process, sugar is converted to alcohol, thus the
inverted correlation of alcohol to residual sugar.

<h3> What was the strongest relationship you found? </h3>

The relationship of residual sugar to density was most strongly correlated.



# Multivariate Plots Section


```{r echo=FALSE, message = FALSE, warning = FALSE, Multivariate_Plots}

wines$alcohol.factor <- factor(round(wines$alcohol))
ggplot(aes(x=round(residual.sugar), y=density), data = wines)+
  geom_line(aes(color = alcohol.factor), stat = 'summary', fun = 'mean')+
  ylim(0.988,1.004)+
  xlim(0,25)

```

In this plot, I showed the relationship between different levels of alcohol 
with residual sugar and density. The lines were noisy, so I rounded alcohol
to the nearest whole number and made it into a factor. I also rounded residual
sugar in order to get smoother lines.



```{r echo=FALSE, message = FALSE, warning = FALSE}

ggplot(aes(total.sulfur.dioxide, y=alcohol), data = wines)+
  geom_point(aes(color = factor(quality)), size=2)+
  scale_color_brewer(type = 'div', palette = 'Set1')+
  xlim(0, 275)

```

This scatter plot shows the relationship between total sulfur dioxide and
alcohol. Quality is the third factor. It shows that higher quality wines have
higher alcohol while lower ones have higher total sulfur dioxide.


```{r echo=FALSE, message = FALSE, warning = FALSE}

ggplot(aes(x=alcohol.factor, y=chlorides), data = wines)+
  geom_col(aes(color = factor(quality)))+
  scale_color_brewer(palette="Set1")
  
```

This bar graph shows that wines with higher chlorides generally had lower
alcohol percentage. It also shows the levels of quality.


# Multivariate Analysis

<h3> Talk about some of the relationships you observed in this part of the 
investigation. Were there features that strengthened each other in terms of 
looking at your feature(s) of interest? </h3>

I was able to further study the relationship between residual sugar, alcohol,
and density. Most wines with higher levels of alcohol appear to have less sugar
and are less dense.

<h3> Were there any interesting or surprising interactions between features?
</h3>

Surprisingly wines lower in alcohol were saltier or higher in chlorides and
lower in quality.

<h3>OPTIONAL: Did you create any models with your dataset? Discuss the
strengths and limitations of your model. </h3>

I felt that modeling the data was a little beyond the scope of the material.
Some of the factors had a middle range where levels were ideal.

------

# Final Plots and Summary


### Plot One
```{r echo=FALSE, message = FALSE, warning = FALSE, Plot_One}

wines <- read.csv('wineQualityWhites.csv')
ggplot(aes(x = quality, y = alcohol), data = wines)+
  geom_point(alpha = 0.1, color = 'blue')+
  geom_smooth(color = 'red', method = "lm", se = FALSE)+
  labs(x = 'Quality Rating', y = 'Alcohol Percentage by Volume')+
  ggtitle('White Wine Quality and Alcohol Content')+
  scale_x_continuous(breaks = c(3,4,5,6,7,8,9))
  
```

### Description One

In this visualization, I show that the quality rating of white wines
is strongly correlated with alcohol content. Because quality is our key factor,
it is important to know this correlation. I used alpha to add some
transparency to make it easier to see the less dense areas. I also added 
a line of regression to clearly show the relationship.

### Plot Two
```{r echo=FALSE, message = FALSE, Plot_Two}
# I sorted by total sulfur dioxide into buckets
wines$total.sulfur.dioxide.bucket <- cut(wines$total.sulfur.dioxide,
                                         c(0,50,100,150,200,250,300))

ggplot(aes(x=round(total.sulfur.dioxide/5) * 5, y=quality), data=subset(wines,                                                           !is.na(total.sulfur.dioxide.bucket)))+
  geom_line(stat = 'summary', fun = 'mean', linetype = 1)+
  geom_smooth()+
  xlim(0,300)+
  labs(y = 'Quality Rating', x = 'Total Sulfur Dioxide mg/L')+
  ggtitle('Quality Rating of White Wines by Total Sulfur Dioxide')+
  scale_x_continuous(breaks = c(0,50,100,150,200,250,300))

```

### Description Two

Here I wanted to show that some factors have middle range where the quality
rating would be highest. I chose total sulfur dioxide because I suspected at
some level, the taste of sulfur would have an undesirable effect on quality.
I sorted the wines into bins by the total sulfur dioxide factor. Next I 
plotted a line graph by plotting the conditional means of the quality factor for
each bin. I had a lot of noise, so I cleaned it by making fewer points by 
dividing and rounding the x factor. I trimmed the x axis at 300 because of some 
outliers. Lastly I used Geom_Smooth to show a nice smooth curve.

### Plot Three
```{r echo=FALSE, warning = FALSE, message = FALSE, Plot_Three}

wines$alcohol.factor <- factor(round(wines$alcohol))
ggplot(aes(x=round(residual.sugar), y=density), data = wines)+
  geom_line(aes(color = alcohol.factor), stat = 'summary', fun = 'mean')+
  labs(x='Residual Sugar mg/L', y=expression(paste('Density g/cm'^'3')))+
  ggtitle('Density of White Wines by Sugar and Alcohol Content')+
  ylim(0.988,1.004)+
  xlim(0,25)+
  labs(color = 'Alcohol %')

```

### Description Three

I wanted to show the three-way relationship between residual sugar, alcohol,
and density. I rounded the alcohol levels and made a factor. I plotted a line
graph that gives a colored line for each alcohol level. I reduced the noise by
rounding the residual sugar factor. Finally I trimmed the x and y axes to get a 
closer look.

# Reflection

The white wine data set that I used was pretty large, nearly 5000 observations.
It gave me a good variety of data with a huge amount of average rated
wines and many outliers. The biggest flaw I found was in the the main
factor, quality. It is based on a rating from 1 to 10 by a median of just three
evaluations by wine experts. That is really small number and no other
information about the wine experts is given.</br>

It was interesting to find that different chemical components had relationships
with the density of the wine. I also found that some factors have a sweet
spot where their levels were optimal for higher quality. There is a clear
relationship between alcohol level and quality.</br>

I struggled to find relationships between many of the factors. They seemed
irrelevant to my analysis. They contributed very little, if not at all, to the
main Quality factor. Overall, I think I was successful at discovering the 
factors that made the biggest difference.</br>

It would be interesting to contrast the white wines with the sister data set of
red wines. I'm curious to see the chemical differences there are between them.
Would red wines generally be higher in density because they are sweeter?</br>
</br>
</br>

